---
title: "Air Quality Analysis"
output: html_document
date: "2026-08-30"
---

# Introduction

## Project Overview

Air quality is an important environmental factor that can affect human health and the environment. This project analyzes air quality measurements using R and the built-in `airquality` dataset.

The project follows a complete data science workflow:

**Data Understanding → Data Quality → Data Cleaning → Exploratory Data Analysis → Statistical Analysis → Time Analysis → Anomaly Detection → Feature Engineering → Modeling → Model Evaluation → Conclusion**

## Project Objective

The main objective of this project is to investigate the relationship between ozone concentration and environmental variables such as temperature, wind speed, and solar radiation.

The project also aims to:

- identify missing values and data-quality issues;

- clean the dataset;

- explore important patterns and relationships;

- identify unusual ozone observations;

- create meaningful features;

- develop prediction models;

- compare model performance using evaluation metrics.

# Dataset

The project uses R's built-in `airquality` dataset.

The dataset contains daily air-quality measurements collected in New York during May to September 1973.

## Variables

| Variable | Description          | Type    |
|----------|----------------------|---------|
| Ozone    | Ozone concentration  | Numeric |
| Solar.R  | Solar radiation      | Numeric |
| Wind     | Wind speed           | Numeric |
| Temp     | Temperature          | Numeric |
| Month    | Month of observation | Integer |
| Day      | Day of observation   | Integer |

## Loading the Dataset

```{r}
data("airquality")

head(airquality)
```

## Data Understanding

```{r pressure, echo=FALSE}
dim(airquality)
```

The dataset contains **153 observations and 6 variables**.

## Column Names

```{r}
names(airquality)

```

## Dataset Structure

```{r}
str(airquality)
```

## Summary Statistics

```{r}
summary(airquality)

```

## Initial Observations

- The dataset contains measurements for ozone, solar radiation, wind, temperature, month, and day.

- Ozone and solar radiation contain missing observations.

- Temperature, wind speed, month, and day are available as numerical variables.

# Data Quality

Before performing analysis, the quality of the dataset was investigated.

Missing Values

```{r}

colSums(is.na(airquality))
```

The original dataset contains:

- 37 missing values in Ozone

- 7 missing values in Solar.R

## Complete Cases

```{r}
sum(!complete.cases(airquality))
```

There are 42 observations containing at least one missing value.

## Duplicate Observations

```{r}
sum(duplicated(airquality))
```

Duplicate observations were checked before cleaning the data.

## Data Quality Observation

Missing values were found mainly in the `Ozone` and `Solar.R` variables. Therefore, missing-data handling was required before further analysis.

# Data Cleaning

The raw dataset was preserved, and a separate cleaned dataset was created.

```{r setup, include=FALSE}
airquality_clean <- airquality 
airquality_clean$Ozone[ is.na(airquality_clean$Ozone) ] <- median(airquality_clean$Ozone, na.rm = TRUE) 

airquality_clean$Solar.R[ is.na(airquality_clean$Solar.R) ] <- median(airquality_clean$Solar.R, na.rm = TRUE)
```

The median values used for imputation were:

- Ozone median = **31.5**

- Solar.R median = **205**

## Checking Missing Values After Cleaning

```{r}
colSums(is.na(airquality_clean))
```

After the cleaning process, the dataset contains no missing values.

```{r}
summary(airquality_clean)
```

## Cleaning Observation

Median imputation was used to replace missing values while keeping the observations available for further analysis.

# Exploratory Data Analysis

Exploratory Data Analysis was performed to understand the distribution of ozone and its relationships with environmental variables.

## Distribution of Ozone

```{r}
hist(airquality_clean$Ozone,
     breaks = 10,
     main = "DISTRIBUTION OF OZONE",
     xlab = "Ozone (ppb)",
     ylab = "Frequency",
     col = "darkblue",  )
```

# Observation:

The distribution of Ozone appears right-skewed.

Most Ozone values are concentrated around 0 to 50 ppb.

The histogram does not clearly show the presence of extreme values;

potential outliers will be investigated using a boxplot.

# **Ozone Boxplot**

```{r}
boxplot(airquality_clean$Ozone,
        main = "Check potential outliers",
        xlab = NULL,
        ylab = "Ozone (ppb)",
        col = "darkred")
```

# Observation:

The median Ozone value is approximately 30 to 35 ppb.

There are potential outliers roughly between 100 and 150 ppb.

The potential outliers are mainly on the right side of the distribution.

## Ozone vs Temperature

```{r}
plot(airquality_clean$Temp,
     airquality_clean$Ozone,
     main = "Relationship between Ozone & Temperature",
     xlab = "Temperature",
     ylab = "Ozone",
     col = "darkgreen",
     lwd = 2
)
```

# Observation:

There appears to be a positive relationship between Temperature and Ozone.

As Temperature increases, Ozone generally tends to increase.

The relationship is not perfectly linear because the points are widely scattered.

A few observations have unusually high Ozone values at higher temperatures.

# Ozone vs Wind

```{r}

plot(airquality_clean$Wind,
     airquality_clean$Ozone,
     main = "Relattonship between Wind & Ozone",
     xlab = "wind (mph)",
     ylab = "Ozone ",
     col = "hotpink",
     lwd = 2
)

```

# Observation:

There appears to be a negative relationship between Wind and Ozone.

As Wind Speed increases, Ozone generally tends to decrease.

The relationship appears fairly noticeable, although the points are scattered.

Some unusually high Ozone values occur at relatively low wind speeds.

# Ozone vs Solar Radiation

```{r}
plot(airquality_clean$Solar.R,
     airquality_clean$Ozone,
     main = "Relationship between Solar Radiation & Ozone",
     xlab = "solar radiation",
     ylab = "Ozone",
     col = "tomato",
     lwd = 2
     
)
```

# Observation:

There appears to be a positive relationship between Solar Radiation and Ozone.

As Solar Radiation increases, Ozone generally tends to increase.

The relationship is fairly scattered, so it does not appear to be very strong.

Some unusually high Ozone values occur at higher Solar Radiation levels.

# Correlation Analysis

Correlation was used to measure the strength and direction of relationships between ozone and the environmental variables.

```{r}
cor(airquality_clean$Ozone, airquality_clean$Temp)
cor(airquality_clean$Ozone, airquality_clean$Wind)
cor(airquality_clean$Solar.R, airquality_clean$Ozone)

```

# Observation:

Temperature has a moderate positive correlation with Ozone (r = 0.6008).

Wind has a moderate negative correlation with Ozone (r = -0.5319).

Solar Radiation has a weak positive correlation with Ozone (r = 0.2955).

These correlation results support the patterns observed

in the scatter plots.

The calculated correlations were:

| Variables               | Correlation |
|-------------------------|-------------|
| Ozone – Temperature     | 0.6008      |
| Ozone – Wind            | -0.5319     |
| Ozone – Solar Radiation | 0.2955      |

## Observation

Temperature has the strongest positive correlation with ozone among the three variables.

Wind has a moderate negative relationship with ozone.

Solar radiation has a weaker positive relationship with ozone.

# Statistical Analysis

## Simple Linear Regression

Simple regression was initially used to investigate individual relationships.

## Ozone and Temperature

```{r}
model_temp <- lm(Ozone ~ Temp,data = airquality_clean)
model_temp

summary(model_temp)

```

# Observation:

Temperature has a statistically significant positive relationship

with Ozone (p \< 2e-16).

The regression coefficient for Temperature is 1.8443,

meaning a 1°F increase in temperature is associated with an

estimated 1.8443 ppb increase in Ozone.

The model explains approximately 36.1% of the variation in Ozone (R² = 0.361).

## Ozone and Wind

```{r}
model_Wind <- lm(Ozone ~ Wind,data = airquality_clean)
model_Wind

summary(model_Wind)

```

# Observation:

Wind has a significant negative relationship with Ozone

(coefficient = -4.3866, p \< 0.05).

For every 1-unit increase in Wind, the model predicts

an average decrease of approximately 4.39 ppb in Ozone.

The model explains approximately 28.29% of the variation

in Ozone (R² = 0.2829).

## Ozone and Solar.R

```{r}

model_Solar.R <- lm(Ozone ~ Solar.R,data = airquality_clean)
model_Solar.R

summary(model_Solar.R)

```

# Observation:

Solar Radiation has a significant positive relationship with Ozone

(coefficient = 0.0975, p \< 0.05).

For every 1-unit increase in Solar.R, the model predicts

an average increase of approximately 0.0975 ppb in Ozone.

The model explains approximately 8.73% of the variation

in Ozone (R² = 0.0873).

## Multiple Linear Regression

```{r}

model_mul_regression <- lm(
  Ozone ~ Temp + Wind + Solar.R,
  data = airquality_clean
)

model_mul_regression

summary(model_mul_regression)
```

# Multiple Linear Regression Observation:

Temperature, Wind, and Solar.R all have statistically significant

relationships with Ozone when considered together (p \< 0.05).

Temperature has a positive coefficient (1.23295), while Wind has

a negative coefficient (-2.78709) and Solar.R has a positive

coefficient (0.05696).

The model explains approximately 47.21% of the variation in Ozone

(R² = 0.4721).

The overall regression model is statistically significant

(F-test p-value \< 2.2e-16).

The model coefficients were approximately:

| Variable        | Coefficient |
|-----------------|-------------|
| Intercept       | -39.35406   |
| Temperature     | 1.23295     |
| Wind            | -2.78709    |
| Solar Radiation | 0.05696     |

The R² value was approximately **0.4721**.

The predictors were statistically significant in the model.

## Interpretation

The temperature coefficient is positive, indicating that ozone tends to increase as temperature increases when the other predictors are held constant.

The wind coefficient is negative, indicating that higher wind speed is associated with lower ozone concentration.

The solar-radiation coefficient is positive, indicating a positive association with ozone.

The model explains approximately **47.21% of the variation in ozone concentration**.

# Time Analysis

The dataset contains `Month` and `Day`, allowing ozone observations to be examined over time.

```{r}
ozone_ts <- ts(airquality_clean$Ozone,
               start = c(1973, 1),
               frequency = 365)

ozone_ts

start(ozone_ts)
end(ozone_ts)
frequency(ozone_ts)
```

# Observation:

The Ozone values were converted into a time-series object using ts().

Since the data contains daily observations, a frequency of 365 was used

to represent the daily time structure of the year 1973.

The time series contains 153 observations, covering the available

May–September 1973 data.

# Plot ozone_ts

```{r}
plot(ozone_ts,
     main = "Daily Obserbations of Ozone",
     xlab = "Evryday Obserbation",
     ylab = "Ozone (ppd)",
     col = "tomato",
     lwd = 2)

```

# Observation:

The time-series plot shows that Ozone varies considerably across

the observation period. Most values are relatively low, but several

high spikes occur, with the highest value around 168 ppb.

No clear continuous increasing or decreasing trend is visible.

Ozone fluctuates considerably, with higher values appearing in

some middle portions of the observation period.

# Moving Average

A moving average was used to smooth short-term fluctuations and make the overall pattern easier to observe.

```{r}
moving_average <- stats::filter(ozone_ts,rep(1/7,7),sides = 2)
moving_average
```

# Observation:

The first and last three observations contain NA because a 7-day

centered moving average requires three observations before and

three observations after each point. These observations are not

available at the beginning and end of the dataset.

```{r}
plot(ozone_ts,
     main = "Ozone with 7-Day Moving Average",
     xlab = "Days",
     ylab = "Ozone (ppb)",
     col = "tomato",
     lwd = 1)

lines(moving_average,
      col = "skyblue",
      lwd = 2)

```

# Observation:

The 7-day moving average smooths the daily fluctuations in Ozone.

Compared with the original Ozone values, the moving average makes

the broader pattern easier to see, while reducing the effect of

short-term spikes.

# Anomaly Detection

Anomaly detection was performed to identify ozone observations that are unusually high or low.

## IQR Method

```{r}
# Find Q1 and Q3
Q1 <- quantile(airquality_clean$Ozone, 0.25)
Q3 <- quantile(airquality_clean$Ozone, 0.75)

# Calculate IQR
IQR_Ozone <- Q3 - Q1

IQR_Ozone

# Lower and upper boundaries
lower_bound <- Q1 - 1.5 * IQR_Ozone
upper_bound <- Q3 + 1.5 * IQR_Ozone

lower_bound
upper_bound
```

The calculated boundaries were approximately:

- Lower boundary = **-16.5**

- Upper boundary = **83.5**

# Detecting Anomalies

```{r}
ozone_anomalies <- airquality_clean[
  airquality_clean$Ozone < lower_bound |
    airquality_clean$Ozone > upper_bound,
]

ozone_anomalies

```

# Observation:

Using the IQR method, 15 Ozone observations were identified as unusually high.

The upper boundary is 83.5 ppb, while the lower boundary is -16.5 ppb.

All detected anomalies are above the upper boundary, ranging from 84 to 168 ppb.

## Visualize the anomalies

```{r}
boxplot(airquality_clean$Ozone,
        main = "Ozone Anomalies",
        ylab = "Ozone (ppb)",
        col = "tomato")

```

# Observation:

The boxplot shows several Ozone observations above the upper whisker.

These points represent unusually high Ozone values and support the

anomalies identified using the IQR method.

```{r}


table(ozone_anomalies$Month)
max(ozone_anomalies$Ozone)
ozone_anomalies[which.max(ozone_anomalies$Ozone), ]
```

# Final Observation:

Most Ozone anomalies occurred in August (7 observations),

followed by July (5 observations). May had 1 anomaly and

September had 2 anomalies. The highest Ozone value was

168 ppb, recorded on August 25.

# Feature Engineering

Feature engineering was used to create additional information from existing variables.

## Create a proper Date

```{r}

dates_text <- paste(1973, airquality_clean$Month,
                    airquality_clean$Day, sep = "-")

class(dates_text)


dates <- as.Date(dates_text)

class(dates)

```

# Observation :

The Month and Day variables were separate, so they were combined

with the year 1973 to create complete date values.

paste() combined Year, Month, and Day into date strings such as

"1973-5-1". These values initially had the character data type.

as.Date() converted the character date strings into actual Date

objects, which can be used for time-based analysis.

## Create Temperature Category

```{r}
airquality_clean$Temp_Category <- cut(
  airquality_clean$Temp,
  breaks = c(-Inf,70,80,Inf),
  labels = c("Low","Medium","High"),
  right = FALSE
)

table(airquality_clean$Temp_Category)
```

# Observation:

Most observations are in the High temperature category (73 observations).

# Create Wind Category

```{r}
airquality_clean$Wind_Category <- cut(
  airquality_clean$Wind,
  breaks = c(-Inf, 10, 15, Inf),
  labels = c("Low Wind", "Medium Wind", "High Wind"),
  right = FALSE
)

table(airquality_clean$Wind_Category)
```

# Observation:

Most observations are in the High temperature category (73 observations).

## Interaction Feature

An interaction feature between temperature and wind was created:

```{r}
airquality_clean$Temp_Wind <- airquality_clean$Temp *
  airquality_clean$Wind



str(airquality_clean)
head(airquality_clean)
summary(airquality_clean)
```

# Observation:

Temp_Wind combines Temperature and Wind into one interaction feature.

It represents how the effect of temperature and wind together may

influence Ozone levels.

# Modeling

The cleaned dataset was divided into training and testing data for prediction.

### Prepare modeling data

```{r}
model_data <- airquality_clean[, c("Ozone", "Temp", "Wind", "Solar.R", "Temp_Wind")]
head(model_data)
colSums(is.na(model_data))

```

## Train/Test Split

```{r}
set.seed(123)

n <- nrow(model_data)

train_index <- sample(1:n, size = 0.8 * n)

train_data <- model_data[train_index, ]
test_data  <- model_data[-train_index, ]

nrow(train_data)
nrow(test_data)
```

# Observation:

The dataset was randomly divided into 80% training data and 20% testing data.

The training data will be used to build the model, while the testing data

will be used later to evaluate its predictions.

## Build the Multiple Linear Regression Model

```{r}

model1 <- lm(Ozone ~ Temp + Wind + Solar.R,
             data = train_data)


summary(model1)
```

# Observation:

The multiple linear regression model explains about 45.73% of the variation

in Ozone. Temperature has a significant positive effect on Ozone, while Wind

has a significant negative effect and Solar.R has a significant positive effect.

Overall, the model is statistically significant (p \< 0.001).

## Make Predictions

```{r}
predictions <- predict(model1, newdata = test_data)
head(predictions)
```

# Observation:

The model generated Ozone predictions for the unseen test data.

The predicted values vary considerably, with some predictions close to

the expected Ozone range while others are negative, such as -14.18.

This shows that the model's predictions will need to be evaluated further.

### Compare Actual vs Predicted Ozone

```{r}
comparison <- data.frame(
  Actual = test_data$Ozone,
  Predicted = predictions
)

head(comparison)
```

# Observation:

The predicted Ozone values are sometimes close to the actual values,

but there are also large differences. For example, an actual value of

7 was predicted as 44.08, and an actual value of 6 was predicted as

-14.18. This indicates that the model does not predict every observation accurately.

## Model 2

```{r}
Ozone ~ Temp + Wind + Solar.R + Temp_Wind

model2 <- lm(Ozone ~ Temp + Wind + Solar.R + Temp_Wind,
             data = train_data)

summary(model2)
```

# Observation:

The improved model explains 56.24% of the variation in Ozone,

increasing the R-squared from 45.73% in model1 to 56.24% in model2.

Temp_Wind is statistically significant (p \< 0.001), indicating that

the interaction between Temperature and Wind contributes meaningfully

to predicting Ozone. Overall, model2 is statistically significant.

# Predictions from Model 1

```{r}
pred_model1 <- predict(model1, newdata = test_data)
```

# Predictions from Model 2

```{r}
pred_model2 <- predict(model2, newdata = test_data)
```

```{r}
model_results <- data.frame(
  Actual = test_data$Ozone,
  Model1_Predicted = pred_model1,
  Model2_Predicted = pred_model2
)

head(model_results)

```

# Observation

Both Model 1 and Model 2 generate predictions that differ from the actual Ozone values.

Model 2 generally gives more reasonable predictions for these test observations.

Model 1 has some large errors, such as predicting -14.18 when the actual value is 6.

Model 2 also has errors, but its predictions are generally closer to the actual values.

Therefore, Model 2 appears to perform better than Model 1 on these sample test observations.

# Model Evaluation Code

| Metric | Complete one-line definition | Better value |
|----|----|----|
| **MAE (Mean Absolute Error)** | Measures the **average absolute difference between the actual Ozone values and the predicted Ozone values**, showing how far predictions are from the real values on average. | **Lower** |
| **RMSE (Root Mean Square Error)** | Measures the **square root of the average squared difference between actual and predicted Ozone values**, giving **more importance to large prediction errors**. | **Lower** |
| **R² (R-squared)** | Measures the **proportion of variation in the actual Ozone values that is explained by the model's predictions**, indicating how well the model fits the data. | **Higher** |

# MAE

```{r}

mae_model1 <- mean(abs(model_results$Actual - model_results$Model1_Predicted))
mae_model2 <- mean(abs(model_results$Actual - model_results$Model2_Predicted))
```

# RMSE

```{r}

rmse_model1 <- sqrt(mean((model_results$Actual - model_results$Model1_Predicted)^2))
rmse_model2 <- sqrt(mean((model_results$Actual - model_results$Model2_Predicted)^2))
```

# R-squared

```{r}

r2_model1 <- 1 - sum((model_results$Actual - model_results$Model1_Predicted)^2) /
  sum((model_results$Actual - mean(model_results$Actual))^2)

r2_model2 <- 1 - sum((model_results$Actual - model_results$Model2_Predicted)^2) /
  sum((model_results$Actual - mean(model_results$Actual))^2)

```

```{r}
model_metrics <- data.frame(
  Model = c("Model 1", "Model 2"),
  MAE = c(mae_model1, mae_model2),
  RMSE = c(rmse_model1, rmse_model2),
  R2 = c(r2_model1, r2_model2)
)
model_metrics
```

# Observation

Model 2 performs better than Model 1 because it has a lower MAE (13.69 vs 14.38)

and lower RMSE (18.90 vs 20.29), indicating smaller prediction errors.

Model 2 also has a higher R² (0.586 vs 0.523), meaning it explains more

variation in Ozone values. Therefore, Model 2 is the better-performing model.

# Compare the Models

```{r}
if (mae_model2 < mae_model1 &&
    rmse_model2 < rmse_model1 &&
    r2_model2 > r2_model1) {
  
  print("Model 2 performs better than Model 1.")
  
} else {
  
  print("Model 1 performs better than Model 2.")
}

```

# Final Observation

Model 2 performs better than Model 1 on the test data.

It has lower MAE and RMSE, indicating smaller prediction errors,

and a higher R², indicating that it explains more variation in Ozone.

Therefore, Model 2 is selected as the better-performing model.

# Final Plot: Actual vs Predicted Ozone

```{r}
plot(
  model_results$Actual,
  model_results$Model1_Predicted,
  main = "Actual vs Predicted Ozone",
  xlab = "Actual Ozone",
  ylab = "Predicted Ozone",
  pch = 19,
  col = "skyblue",
  lwd = 2
)

points(
  model_results$Actual,
  model_results$Model2_Predicted,
  pch = 19,
  col = "tomato",
  lwd = 02
)

abline(0, 1, lty = 2)

legend(
  "topleft",
  legend = c("Model 1", "Model 2", "Perfect Prediction"),
  pch = c(19, 19, NA),
  lty = c(NA, NA, 2),
  col = c("skyblue", "tomato", "black")
)
```

# Final Observation

The Actual vs Predicted plot shows that Model 2's predictions are generally

closer to the perfect prediction line than Model 1.

This visually supports the evaluation results, where Model 2 has lower MAE

and RMSE and higher R². Therefore, Model 2 provides better Ozone predictions

for the test data from the cleaned airquality dataset.

# Final Findings

The major findings from the project are:

1.  The dataset contains 153 observations and 6 original variables.

2.  Missing values were present in Ozone and Solar Radiation.

3.  Median imputation was used to handle missing values.

4.  Temperature showed the strongest positive correlation with ozone.

5.  Wind showed a moderate negative relationship with ozone.

6.  Multiple regression explained approximately 47.21% of ozone variation in the initial statistical model.

7.  Ozone showed substantial fluctuations over the observation period.

8.  Several unusually high ozone observations were detected using the IQR method.

9.  Anomalies were retained rather than automatically removed because they may represent genuine environmental observations.

10. The temperature-wind interaction feature improved model performance.

11. Model 2 achieved lower MAE and RMSE and higher R² than Model 1.

# Conclusion

This project demonstrated a complete data science workflow using R and the `airquality` dataset.

The analysis began with understanding the dataset and identifying data-quality issues. Missing values were handled using median imputation, after which exploratory analysis was performed to investigate ozone distributions and relationships with environmental variables.

Statistical analysis showed that temperature had a strong positive relationship with ozone, while wind had a negative relationship. Multiple linear regression was then used to model ozone concentration using environmental variables.

Time-based analysis was used to investigate ozone fluctuations, while anomaly detection identified unusually high ozone observations. Feature engineering introduced an interaction between temperature and wind.

Finally, two predictive models were compared using MAE, RMSE, and R². The second model performed better, demonstrating that the engineered interaction feature provided useful additional information for ozone prediction.

Overall, the project demonstrates how R can be used to move from raw environmental data to data cleaning, analysis, interpretation, feature engineering, and predictive modeling.
